Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process

### 2023-09-15 report/appendix_h.Rmd
Add first rough draft

We used LOO-CV to compare the predictive performance of the two models for each data generating process. The model comparisons can be found for each data generating process in Table \ref{tab:loossim}. The results rule out the possibility that the mixture model does always lead to higher predictive performance. Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a uni-modal process. However, for the data generated with a bi-modal process, the mixture model shows a substantially higher predictive performance. In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence [@sivula2020uncertainty], shows that the mixture model performs 11.6 times better than the uni-modal model. In comparison, for the uni-modal data, the uni-modal model performs only 0.77 times better than the bi-modal mixture model. Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the uni-modal is preferred by the law of parsimony.

>>>>

We used LOO-CV to compare the predictive performance of the two models for each data generating process. The model comparisons can be found for each data generating process in Table \ref{tab:loossim}. The results rule out the possibility that the mixture model does always lead to higher predictive performance. Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a uni-modal process. However, for the data generated with a bi-modal process, the mixture model shows a substantially higher predictive performance. In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence (Sivula et al., 2020), shows that the mixture model performs 11.6 times better than the uni-modal model. In comparison, for the uni-modal data, the uni-modal model performs only 0.77 times better than the bi-modal mixture model. Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the uni-modal is preferred by the law of parsimony.

### 2023-09-15 report/appendix_h.tex
Add first rough draft

## Simulation

@@

This model is largely identical to the models before but reduced to its main parameters (but not mixed effects for participants). The model includes two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.

@@

Again, this model is a simplified version of the uni-modal models used in the main analysis. The model assumes a log-Gaussian distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.

@@

```{r simdata, fig.cap= "Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
files <- list.files("../mixture-model-sim/data", full.names = T)

map_dfr(files, ~read_csv(.) %>% 
        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
  mutate(across(Dataset, ~recode(., mogdata = "bi-modal", lmdata = "uni-modal"))) %>% 
  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
  geom_density(alpha = .25) +
  scale_colour_colorblind() + 
  scale_fill_colorblind() +
  scale_x_log10(labels = ~format(., big.mark = ",")) +
  theme(text = element_text(size = 10)) +
  labs(colour = "Data process", 
       fill = "Data process", 
       x = "y", 
       y = "Density")
```

For each of these two data sets we simulated 1,000 observations. We run 2 models -- a bi-modal mixture model and a uni-modal model -- each for both data sets. Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup. Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}. The parameters are shown by type of data generating process along with the true parameter values. Parameter value estimates are shown by type of Bayesian model. The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process. Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they uni-modal model.

```{r}
mog1 <- read_csv("../mixture-model-sim/stanout/mog_mogdata.csv") %>% 
  mutate(data = "mog")
mog2 <- read_csv("../mixture-model-sim/stanout/mog_lmdata.csv") %>% 
  mutate(data = "lm")
lm1 <- read_csv("../mixture-model-sim/stanout/lm_lmdata.csv") %>% 
  mutate(data = "lm")
lm2 <- read_csv("../mixture-model-sim/stanout/lm_mogdata.csv") %>% 
  mutate(data = "mog")

mog_sim <- bind_rows(mog1, mog2) %>% 
  summarise(across(value, list(est = mean,
                               lo = lower,
                               up = upper),
                   .names = "{.fn}"),
            .by = c(name, data)) %>% 
  mutate(across(where(is.numeric), ~round(., 2)),
         across(where(is.numeric), ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
         across(est, ~str_c(., " [", lo, ", ", up, "]")),
         across(name, ~str_c("\\", .)),
         across(name, ~str_replace(., "a_e", "a^2_1")),
         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>%
  select(-lo, -up) %>% 
  pivot_wider(names_from = data, values_from = est) %>% 
  mutate(true = c(5, 1, .35, .25, .5),
         across(true, ~ifelse(str_detect(name, "prob"), dezero(., 2), .)),
         across(name, ~str_replace(., "prob", "theta"))) %>% 
  relocate(name, true) %>% 
  unite("name", name:true, sep = " = ")

lm_sim <- bind_rows(lm1, lm2) %>% 
  summarise(across(value, list(est = mean,
                               lo = lower,
                               up = upper),
                   .names = "{.fn}"),
            .by = c(name, data)) %>% 
  mutate(across(where(is.numeric), ~round(., 2)),
         across(est, ~str_c(., " [", lo, ", ", up, "]")),
         across(name, ~str_c("\\", .)),
         across(name, ~str_replace(., "a_e", "a^2_1")),
         across(name, ~str_replace(., "ap_e", "a^2_2"))) %>% 
  select(-lo, -up) %>%
  pivot_wider(names_from = data, values_from = est) %>% 
  mutate(true = as.character(c(5, .25))) %>% 
  relocate(name, true) %>% 
  unite("name", name:true, sep = " = ")

```

```{r simparam, results='asis'}
table <- bind_rows(mog_sim %>% mutate(model = "mog"),
                   lm_sim %>% mutate(model = "lm")) %>% 
  relocate(model) %>% 
  mutate(across(name, ~str_c("$", ., "$")))

names(table) <- c("Model", "Parameter with true value", "Bi-modal model", "Uni-modal model")

apa_table(table[,-1], 
          escape = FALSE,
      col_spanners = list("Estimate with 95\\% PI" = 2:3),
      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
      align = c("l", rep("r", 2)),
      stub_indents = list(`Bi-modal data` = 1:5, 
                          `Uni-modal data` = 6:7))

```

We used LOO-CV to compare the predictive performance of the two models for each data generating process. The model comparisons can be found for each data generating process in Table \ref{tab:loossim}. The results rule out the possibility that the mixture model does always lead to higher predictive performance. Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a uni-modal process. However, for the data generated with a bi-modal process, the mixture model shows a substantially higher predictive performance. In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence (Sivula et al., 2020), shows that the mixture model performs 11.6 times better than the uni-modal model. In comparison, for the uni-modal data, the uni-modal model performs only 0.77 times better than the bi-modal mixture model. Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the uni-modal is preferred by the law of parsimony.

```{r loossim, results='asis'}
table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
  select(1:6) %>% 
  mutate(
#    elpd_se_ratio = abs(elpd_diff / se_diff),
    across(ends_with("diff"), ~round(., 1)),
    across(ends_with("loo"), ~round(., 0)),
    across(where(is.numeric), ~format(., big.mark = ",")),
    across(everything(), ~as.character(.)),
    across(everything(), ~str_trim(.)),
    across(everything(), ~str_replace_all(., "NaN", "--")),
    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
    across(c(data, model), ~recode(., mog = "Bi-modal mixture model",
                                      lm = "Uni-modal model"))) %>% 
  select(data, model, starts_with("elpd"))

names(table) <- c("Data", "Model", "$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$")

apa_table(table[,-1], 
          escape = FALSE,
      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
      align =c("l", rep("r", 2)),
      stub_indents = list(`Data: Bi-modal mixture process` = 1:2, 
                          `Data: Uni-modal process` = 3:4))

```

>>>>

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{simulation}{%
\subsection{Simulation}\label{simulation}}

@@

This model is largely identical to the models before but reduced to its
main parameters (but not mixed effects for participants). The model
includes two log-normal distributions with a mixing proportion
\(\theta\) of which the distribution of shorter values has a mean of
\(\beta\) and a standard deviation \(\sigma^2_2\); the second
distribution of longer values is constrained to have a mean that is
larger by a factor of \(\delta\) and has a larger standard deviation
\(\sigma^2_1\).

@@

Again, this model is a simplified version of the uni-modal models used
in the main analysis. The model assumes a log-Gaussian distribution with
a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).

@@

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/simdata-1}

}

\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}(\#fig:simdata)
\end{figure}

For each of these two data sets we simulated 1,000 observations. We run
2 models -- a bi-modal mixture model and a uni-modal model -- each for
both data sets. Models were run with 3 chains, with each 6,000
iterations of which 3,000 were warmup. Estimates with 95\% probability
intervals are shown in Table \ref{tab:simparam}. The parameters are
shown by type of data generating process along with the true parameter
values. Parameter value estimates are shown by type of Bayesian model.
The results show that either of the two Bayesian models succesfully
uncovered the model parameters of the data with its corresponding data
generating process, as shown in Table \ref{tab:simparam}, but less so
when the model was applied to data generated with the incorrect
underlying process. Particularly the mixing proportion \(\theta\) and
the slowdown parameter \(\delta\) were not uncovered at all by they
uni-modal model.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:simparam}Uncovered parameter estimates with 95\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.}

\begin{tabular}{lrr}
\toprule
 & \multicolumn{2}{c}{Estimate with 95\% PI} \\
\cmidrule(r){2-3}
Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
\midrule
Bi-modal data &  & \\
\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
Uni-modal data &  & \\
\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

We used LOO-CV to compare the predictive performance of the two models
for each data generating process. The model comparisons can be found for
each data generating process in Table \ref{tab:loossim}. The results
rule out the possibility that the mixture model does always lead to
higher predictive performance. Indeed, the mixture model showed a
slightly lower predictive performance for the data that were generated
with a uni-modal process. However, for the data generated with a
bi-modal process, the mixture model shows a substantially higher
predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
and its standard error, as metric for the strength of evidence (Sivula
et al., 2020), shows that the mixture model performs 11.6 times better
than the uni-modal model. In comparison, for the uni-modal data, the
uni-modal model performs only 0.77 times better than the bi-modal
mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
the model is negligible, the uni-modal is preferred by the law of
parsimony.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:loossim}Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.}

\begin{tabular}{lrr}
\toprule
Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
\midrule
Data: Bi-modal mixture process &  & \\
\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
Data: Uni-modal process &  & \\
\ \ \ Uni-modal model & -- & -5,030 (24)\\
\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
\bottomrule
\addlinespace
\end{tabular}

\begin{tablenotes}[para]
\normalsize{\textit{Note.} $\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\Delta\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.}
\end{tablenotes}

\end{threeparttable}
\end{center}

\end{table}
\end{appendix}

### 2023-09-15 report/manuscript.Rmd
Add first rough draft

documentclass     : "apa7"
classoption       : "man"
output            : 
  papaja::apa6_pdf:
    keep_tex: TRUE
    includes:
      after_body: 
        - "appendix_a.tex"
        - "appendix_b.tex"
        - "appendix_c.tex"
        - "appendix_d.tex"
        - "appendix_e.tex"
        - "appendix_g.tex"
        - "appendix_f.tex"
        - "appendix_g.tex"

@@

Interestingly even the youngest sample in our data pool (C2L1)

showed evidence that pauses at before sentence locations were not obligatory and similar to pauses before words. This is interesting because previous research suggested that encoding of subsequent planning units is more separated -- and therefore operates in a serial fashion -- in inexperienced writers [@olive2014toward].

@@

Using cross-validation and a simulation we ruled out that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting. Modelling details can be found in Appendix \ref{simulation}.

@@

All plantra results are L2 but not the other studies

to what extent do pauses reflect reading behaviour?

Say how the results don't contradic the reported studies but speaks better to their conclusions 
but the modelling frameword doesn't require us to impose pause thresholds

(2) dichotomising latencies into pauses and non-pauses discards much of the information that is captured within a keystroke log.

One strategy for handling the complex distribution of latency from spontaneous text production is to avoid reporting central tendency altogether and to just count latencies that exceed a specific threshold – traditionally set at 2 seconds. This practice, inherited from early research in speech production (see Rochester, 1973 for a review) has been widely adopted by writing researchers, with studies dating back to at least the mid 1990s (Foulin, 1995, 1998). Pause counting has two disadvantages. First, researchers have to make an a priori decision about where to set the pause threshold. Current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision. So while a 2 s threshold undoubtedly captures an interesting distinction – processing that occurs the range zero to 2 s is very likely to be qualitatively different from processing that takes more than 2 s. However the same could be argued for any threshold between perhaps 250 ms and 10 that a researcher might care to choose (Chenu et al., 2014). Moreover, as we have already discussed, interpretation of latencies depends on location within the text. A sentence-initial pause of 2 s has a very different interpretation from a 2 s pause that occurs before or within a word. A second disadvantage of pause counting is that dichotomizing latencies discards much of the information that is captured within a keystroke log or digitized handwriting trace. We therefore require statistical methods that model the variance across the full range keystroke latencies (or similar measures from handwriting).

Open: to what extent are even short latencies longer at sentences. This is obvious for when multiple keystrokes are summed across at sentences boundaries as there are typically three key transitions necessary to start a sentence: (1) .^_ , (2) _^[shift], (3) [shift]^C
We believe the second one is the linguistically meaningful one
I think give Davids papers as examples, if he uses sums instead of means (other studies?)

I just read the abstract of the first one, but the neural oscillations that they describe are occur every 153 ms (6.5 hz). Which is kind of the magic number for the short-duration IKI component.

@@

I think the serial account isn't all inconsistent with the mixture model. I don't think the serial account says that pauses must occure on linguistic edges but between bursts which can in principlpe happen anywhere (but why would they be random). Obviously the problem here is that one needs to postulate what a pause is instead of leaving it to the model to work out. The serial account and the cascading account make very similar predictions for inexperienced writers.

To what extent do pauses reflect reading back. Cite literature from Marks first SPL2 draft on reading during writing.

- Reading is necessary for revision: check if writing goals have been met.
- Planning what to say next: reinstating ideas after interruption of writing process.
- Error evaluation: e.g. spelling, grammar

- we removed transitions that terminate in editiing so lookbacks are not associaed with revisions.

- Adults writing argumentative essays
- @torrance2016reading: 
  - before sentences 45% lookbacks
  - before words 12% lookbacks
  - within word 5% lookbacks 
- 36% sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined]
- Limited work: @alamargot2010using; @beers2010adolescent; @de2018exploring; @van2010reading

>>>>

documentclass     : "apa7"
classoption       : "man"
output            : 
  papaja::apa6_pdf:
    keep_tex: TRUE
    includes:
      after_body: 
        - "appendix_a.tex"
        - "appendix_b.tex"
        - "appendix_c.tex"
        - "appendix_d.tex"
        - "appendix_e.tex"
        - "appendix_g.tex"
        - "appendix_f.tex"
        - "appendix_h.tex"

@@

Interestingly even the youngest sample in our data pool (C2L1) showed evidence of parallel planning. For example, @olive2014toward described a cascading model that is operating in a serial fashion for inexperienced or struggling writers. This is because of task demands reduce the ability to plan in parallel to processing which leads to a separation of planning units that therefore operate in a serial fashion. However, the results presented in this paper suggest that pausing behaviour in young writers largely mirror results from more experienced writers with the notable exception that pause durations are similar before words and sentence indicating a tendency to prepare utterances in small lexically-based planning units possibly smaller than the clause. The main differences for the young writers was their fluent writing execution is substantially slower compared to more experienced writers in the sample. Also for L2 writers (PlanTra, SPL2) we observe pausing patterns that do not resemble a serialised writing process.

@@

A possible concern with our results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit. We addressed this concern by using cross-validation techniques for model comparison which is preventing overfitting models. In addition we used a simulation approach to compare a uni-modal model and a bi-modal mixture model. We simulated two data sets that were either based on a uni-modal or bi-modal random data generating process. In other words, this approach allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value. Modelling details can be found in Appendix \ref{simulation}. As might be expected, we found, that the bi-modal mixture was successful at uncovering the parameter values of the data generated with a bi-modal mixture models; we observed the same for the uni-modal model of the data generated with a uni-modal process. Parameters were not successfully uncovered when we switched model type and data set. Importantly though, cross-validation did not show a higher predictive performance for the mixture model, compared to the uni-modal model,  when applied to the uni-modal data (but for the bi-modal data). Therefore we can rule out the possibility that the overwhelming statistical support for bi-modal models can be explained in terms of model overfitting.

Throughout this paper we assumed that hesitations are indicators of planning upcoming ideas and encoding of linguistic units. Alternatively at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text. There is a limited amount of work that has used eye tracking to investiatge regressive eye movement behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading]. There are at least three reasons why writers look back into their text: (1) reading is necessary for revision (i.e. writers must check if writing goals have been met); (2) using text to cue or reinstating ideas after interruption of writing process; (3) error evaluation: e.g. spelling, grammar. We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions. However, frequencies of lookback during writing follow a similar pattern as the one we observed for pausing: @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined]. This is except, look backs are more frequent before sentences compared to words while we found that pause frequencies appear equally often before sentences and words. Finally two of the data sets in our poor (CATO, GUNNEXP2) included a manipulation where the produced text was masked (or not). In a reanalysis we found no substantial differences in pausing behaviour for the masked condition; see Appendix \ref{masking-effect-cato-gunnexp2}. In other words, even though reading is likely be explain some proportion of our results it is overall rare, certainly in samples of university students.

@@

~~~~~
Say how the results don't contradic the reported studies but speaks better to their conclusions 
but the modelling frameword doesn't require us to impose pause thresholds
~~~~~

~~~~
Move from results to discussion?
Open: to what extent are even short latencies longer at sentences. This is obvious for when multiple keystrokes are summed across at sentences boundaries as there are typically three key transitions necessary to start a sentence: (1) .^_ , (2) _^[shift], (3) [shift]^C
We believe the second one is the linguistically meaningful one
I think give Davids papers as examples, if he uses sums instead of means (other studies?)

~~~

neural oscillationsoccur every 153 ms (6.5 hz). Which is kind of the magic number for the short-duration IKI component.

@@

~~~~
I think the serial account isn't all inconsistent with the mixture model. I don't think the serial account says that pauses must occure on linguistic edges but between bursts which can in principlpe happen anywhere (but why would they be random). Obviously the problem here is that one needs to postulate what a pause is instead of leaving it to the model to work out. The serial account and the cascading account make very similar predictions for inexperienced writers.

Maybe distinguih between a strong version of serial model and a weak version that is basically a chunky cacading model
